---
title: "scMultiSim Examples"
output: html_notebook
---
```{r "setup", include=FALSE}
require("knitr")
opts_chunk$set(fig.width=4, fig.height=3)
```

First, load the package locally:
```{r, message=F, warning=F, eval=T}
(function() {
  installed <- installed.packages()[,"Package"]
  install <- function(list, fn) {
    pkg <- setdiff(list, installed)
    if (length(pkg)) fn(pkg)
  }
  
  r_packages <- c(
    "devtools", "dplyr", "ggplot2", "Rtsne",
    "reshape", "ape", "phytools", "repr", "KernelKnn",
    "gridExtra", "parallel", 'foreach', 'phytools', "doParallel",
    "zeallot", "gtools", "gplots", "stringi", "roxygen2"
  )
  install(r_packages, install.packages)
  
  if (!require("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
    BiocManager::install(version = "3.15")
  }
  bioc_packages <- c('Biobase','IRanges','S4Vectors','SummarizedExperiment')
  install(bioc_packages, BiocManager::install)
})()
```

```{r}
library("devtools")
devtools::load_all(".")
library(gridExtra)
library("scMultiSim")
```

## Simulating True Counts

Simulate true counts by calling `sim_true_counts(options)` where `options` is a
list. Use `scmultisim_help()` to get help.

```{r}
scmultisim_help("options")
```

scMultiSim provides two sample GRNs, `GRN_params_100` and `GRN_params_1139`.
Load them first.
```{r}
data(GRN_params_100)
GRN_params <- GRN_params_100
GRN_params
```

scMultiSim also provides three differentiation trees.
`Phyla5()` and `Phyla3()` return bifurcating trees with 5 and 3 leaves respectively.
`Phyla1()` returns only a single branch, which can be useful when we don't want any trajectory.
```{r}
Phyla5(plotting = T)
Phyla3(plotting = T)

# It's not possible to plot Phyla1() because it only contains 1 branch connecting two nodes.
Phyla1()
```

Now, we create the options list.
The undocumented option `debug` will print out some useful information and return the internal simulation session in the result.
```{r}
options_ = list(
  rand.seed = 0,
  GRN = GRN_params,
  num.cells = 1000,
  num.cifs = 500,
  tree = Phyla5(),
  diff.cif.fraction = 0.8,
  do.velocity = T,
  intrinsic.noise = 1
  # debug = T
)
```

Run the simulation and check what kind of data is in the returned result:
```{r}
results <- sim_true_counts(options_)
names(results)
```

Visualize the true counts and ATAC-seq data using `plot_tsne()`:
```{r}
plot_tsne(log2(results$counts + 1),
          results$cell_meta$pop,
          legend = 'pop', plot.name = 'True RNA Counts Tsne')
plot_tsne(log2(results$atacseq_data + 1),
          results$cell_meta$pop,
          legend = 'pop', plot.name = 'True ATAC-seq Tsne')
```

We can also add the technical variation and batch effect to the true counts:
```{r}
add_expr_noise(results)
divide_batches(results, nbatch = 2)

plot_tsne(log2(results$counts_with_batches + 1),
          results$cell_meta$pop,
          legend = 'pop', plot.name = 'RNA Counts Tsne with Batches')
```

## Simulating cell-cell interaction

The cell-cell interaction works similar to a GRN; supply the `spatial.params` to
simulate spatial cell-cell interactions.
```{r,cache=T}
options_ <- list2(
  GRN = data.frame(
    target = 1:10,
    regulator = 11:20,
    effect = 3
  ),
  num.genes = 50,
  num.cells = 500,
  num.cifs = 500,
  tree = Phyla3(),
  do.velocity = F,
  cci = list(
    params = data.frame(
      target = 21:25,
      regulator = 26:30,
      effect = 3
    ),
    max.neighbors = 2
  )
)

results <- sim_true_counts(options_)
```

Now we plot the cell locations:
```{r fig.width = 6, fig.asp = 1}
plot_cell_loc(results)
```

## Simulating dynamic GRN

First, call the following function to check the usage of dynamic GRN.
```{r}
scmultisim_help("dynamicGRN")
```

Here we use `Phyla1()` as the differentiation tree to remove the effect of the trajectory. Additionally, we can use `discrete.cif = T` to simulate discrete cell population.
```{r}
options_ = list(
  rand.seed = 0,
  GRN = GRN_params,
  num.cells = 1000,
  num.cifs = 500,
  tree = read.tree(text="(t1:10);"),
  diff.cif.fraction = 0.8,
  # discrete.cif = T,
  do.velocity = F,
  dynamic.GRN = list(
    cell.per.step = 3,
    num.changing.edges = 5,
    weight.mean = 0,
    weight.sd = 4
  ),
  intrinsic.noise = 0.5,
  debug = T
)

results <- sim_true_counts(options_)
```

`results$cell_specific_grn` is a list containing the gene effects matrix for each cell. Each row is a target and each column is a regulator. The corresponding gene names are displayed as column and row names.
```{r}
# GRN for cell 1 (first 10 rows)
results$cell_specific_grn[[1]][1:10,]
```

Since we set `cell.per.step = 3`, we expect each adjacent 3 cells share the same GRN:
```{r}
print(all(results$cell_specific_grn[[1]] == results$cell_specific_grn[[2]]))
print(all(results$cell_specific_grn[[2]] == results$cell_specific_grn[[3]]))
print(all(results$cell_specific_grn[[3]] == results$cell_specific_grn[[4]]))
```